---
title: "REPORT"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this part, eQTLQC perfoms quality control and normalization based on Metrics data (TPM/RPKM/FPKM/Readcount).


```{r echo=FALSE, warning=FALSE, results='hide', message=FALSE}
# Load packages
if (!require("rjson")){
  install.packages("rjson")
  library(rjson)
}
if (!require("RColorBrewer")){
  install.packages("RColorBrewer")
  library(RColorBrewer)
}
if (!require("ape")){
  install.packages("ape")
  library(ape)
}
if (!require("reshape2")){
  install.packages("reshape2")
  library(reshape2)
}
if (!require("tidyverse")){
  install.packages("tidyverse")
  library(tidyverse)
}
if (!require("sva")){
  install.packages("sva")
  library(sva)
}
```

## Basic quality control steps
```{r echo=FALSE, warning=FALSE, results='hide', message=FALSE}
# Load Configuration file
json_data<-fromJSON(file="Tool.config")

metaInfo <- toupper(json_data$config$transcriptome$transcriptomeQC$covariates$covariates_tag)=="TRUE"
if(metaInfo){
  metaInfoFile <- json_data$config$transcriptome$transcriptomeQC$covariates$covariates_file
  metaInfo_sample_id_column <- json_data$config$transcriptome$transcriptomeQC$covariates$sample_id_column
  metaInfo_batch_column <- json_data$config$transcriptome$transcriptomeQC$covariates$batch_column
  metaInfo_gender_column <- json_data$config$transcriptome$transcriptomeQC$covariates$gender_column
  metaInfo_adjust_covariates <- json_data$config$transcriptome$transcriptomeQC$covariates$adjust_covariates
}
csvfile <- "TPM_matrix.csv"
RLEFilterPercent <-json_data$config$transcriptome$transcriptomeQC$RLEFilterPercent
DSFilterPercent <- json_data$config$transcriptome$transcriptomeQC$DSFilterPercent
pvalues.cut <- json_data$config$transcriptome$transcriptomeQC$hcluster$pvalues.cut
topk_genes <- json_data$config$transcriptome$transcriptomeQC$hcluster$topk_genes
cluster_percent <- json_data$config$transcriptome$transcriptomeQC$hcluster$cluster_percent
treesNum <- json_data$config$transcriptome$transcriptomeQC$hcluster$cluster_level
low_expr_TPM <- json_data$config$transcriptome$transcriptomeQC$low_expr_TPM
low_expr_TPM_percent <- json_data$config$transcriptome$transcriptomeQC$low_expr_TPM_percent
pseudo_count <- json_data$config$transcriptome$transcriptomeQC$TPM_pseudo_count
Female_specific_gene <- json_data$config$transcriptome$transcriptomeQC$gender_check$Female_specific_gene
Male_specific_gene<- json_data$config$transcriptome$transcriptomeQC$gender_check$Male_specific_gene
print("Configuration file load successfully!")
```


```{r echo=FALSE, warning=FALSE, results='hide'}
if(metaInfo)
{
  Meta <- read.table(metaInfoFile, header = TRUE, stringsAsFactors = TRUE)
  print("covariates file loaded sucessfully!")
  adjust_covariates = metaInfo_adjust_covariates %>% str_split(pattern = ",\\s") %>% unlist
  if(!(metaInfo_sample_id_column %in% names(Meta))){
    print(paste0("Cannot detect sample ID column: ", metaInfo_sample_id_column, " , Exit!"))
    quit(save = "no", status = 1, runLast = FALSE)
  }else{
    if(length(unique(Meta[[metaInfo_sample_id_column]])) != nrow(Meta)){
      print(paste0("Sample ID column: ", metaInfo_sample_id_column, ", has duplicated IDs. Please remove duplicates, Exit!"))
      quit(save = "no", status = 1, runLast = FALSE)
    }
  } # sample_id is mandatory
  if(sum(adjust_covariates %in% names(Meta)) != length(adjust_covariates)){
    print(paste0("Cannot detect covariates column: ", paste(adjust_covariates[!(adjust_covariates %in% names(Meta))], collapse= ", "), " ,Exit!"))
    quit(save = "no", status = 1, runLast = FALSE)
  } # covariates to be adjusted should have accurate names
  if((metaInfo_batch_column!="") && (metaInfo_batch_column %in% adjust_covariates)){
    print(paste0("Covariate ", metaInfo_batch_column, " detected. We will adjust batch effects."))
    batch_adjust_signal <- TRUE
    adjust_covariates <- adjust_covariates[adjust_covariates != metaInfo_batch_column] # eQTLQC adjust batch effects separately.
  }else{
    batch_adjust_signal <- FALSE
  }
  if((metaInfo_gender_column!="") && (metaInfo_gender_column %in% adjust_covariates)){
    print(paste0("Covariate ", metaInfo_gender_column, " detected. We will check gender-mismatched samples."))
    gender_adjust_signal <- TRUE
  }else{
    gender_adjust_signal <- FALSE
  }
  rownames(Meta) <- Meta[[metaInfo_sample_id_column]]
}
```

```{r echo=FALSE, warning=FALSE, results='hide'}
# Load expression data, exclude low-expressed genes, samples without meta information
TPM_data <- read.csv(csvfile,header = TRUE) # The input file should has sample ID as column name; the first column is gene ID 
names(TPM_data)[1] <- "feature"
if( sum(duplicated(TPM_data$feature)) > 0){
  print("Feature ID (e.g. gene names) should be in the first column. Please remove duplicated feature IDs, Exit!")
  quit(save = "no", status = 1, runLast = FALSE)
}else{
  rownames(TPM_data) <- TPM_data$feature
  TPM_data = TPM_data[,-1]
  loaded_sample_count <- ncol(TPM_data)
}
# make sure every expr column is in numeric type
matrix_check <- map(TPM_data, is.numeric) %>% unlist
if(sum(!matrix_check) > 0){
  print("The following column(s) in expression matrix is/are NOT in numeric type. Plese check, Exit!")
  print(paste(names(matrix_check)[!matrix_check], collapse = "; "))
  quit(save = "no", status = 1, runLast = FALSE)
}
print("Gene expression profiles loaded successfully!")
print(paste(nrow(TPM_data), "genes and", ncol(TPM_data), "samples are loaded from", csvfile, sep = " "))

idx <- intersect(colnames(TPM_data),rownames(Meta))
Meta <- Meta[idx,]
TPM_data <- TPM_data[,idx]
if(loaded_sample_count-length(idx) >0){
  print(paste(loaded_sample_count-length(idx), "samples are excluded because of sample mismatching between covariates and expression data.")) 
}

```

```{r, echo=FALSE, warning=FALSE}
if(metaInfo && gender_adjust_signal && (Female_specific_gene!='')){
   logtpm = log10(TPM_data + pseudo_count)
   print("Generate gender specific genes' expression plot:")
   print(paste("Female specific gene:", Female_specific_gene, "and male specific gene:", Male_specific_gene, "will be used as classification features.", sep = " "))
   if(Female_specific_gene %in% rownames(TPM_data) && Male_specific_gene %in% rownames(TPM_data)){
    # Gender mismatch
      if (!require("e1071")){
      install.packages("e1071")
      library(e1071)
     }
    mismatchFilterList = list()
    chrX=Female_specific_gene #'ENSG00000229807'  # XIST
    chrY=Male_specific_gene   #'ENSG00000129824'  # RPS4Y1
    d=as.data.frame(t(logtpm[c(chrX,chrY),])); 
    colnames(d) <- c(Female_specific_gene, Male_specific_gene)
    d$Sex = Meta[match(rownames(d), rownames(Meta)), metaInfo_gender_column]
    if(length(unique(d$Sex))<2){
      print("Less than 2 sex types, please check the sex covariate. Skip!")
    }else{
    plot(d[[Female_specific_gene]], d[[Male_specific_gene]], xlab= paste0("Log expression of ", Female_specific_gene), ylab=paste0("Log expression of ", Male_specific_gene), col= 'white',bg=ifelse(tolower(d$Sex)=="female",'red','blue'), pch=21, bty="n", main="Gender-specific gene expression")
    legend('bottomleft',pch=21,c("Female","Male"), col='white',pt.bg=c("red","blue"), bty='n')
    model1 = svm(Sex ~ .,data = d,type = 'C-classification')
    plot(model1,d)
    pSex = predict(model1)
    matchlist = (pSex == d$Sex)
    mismatchList = which(matchlist==FALSE)
    clusterFilterList = rownames(d)[mismatchList]
    clusterFilterList <- as.character(clusterFilterList)
    if(length(clusterFilterList) > 0){
      print("The following samples are gender-mismatched, and have been excluded:")
      print(paste(clusterFilterList, collapse = "; "))
    }else{
      print("All samples in expression data have correct gender information.")
    }
    }
    
   }else{
     print(paste("Cannot detect at least one gene of:", Female_specific_gene, "and", Male_specific_gene, sep = " "))
   }
   }else{
  print("Cannot detect `sex` covariate, skip gender-mismatch checking.")
}
```



```{r echo=FALSE, warning=FALSE, results='hide'}
keep_genes_idx <- (rowMeans(TPM_data>low_expr_TPM)>low_expr_TPM_percent) 
TPM_data = TPM_data[keep_genes_idx,]
print(paste(sum(1 - keep_genes_idx), "genes are filtered, because >", low_expr_TPM_percent*100, "% samples have expression values <", low_expr_TPM))
print(paste(sum(keep_genes_idx), "genes left."))
clusterFilterList <- c()
logtpm = log10(TPM_data + pseudo_count)
```


## Identify sample outliers with abnormal expression profiles using three statistical methods

### Relative Log Expression (RLE) Analysis

RLE (Relative Log Expression) analysis assumes that, most genes' expression values should be stable across samples, and only a few genes have differential expression levels. Given a gene expression matrix *G* with gene in row and sample in column, RLE analysis has the following steps: (1) for each gene *i* , calculate its median expression level across *N* sample as *Medi* ; (2) for each sample *j* , and each gene *i*, substract the median expression level of each gene (*Medi*) from original expression values *G(i,j)*; (3) For each sample, the residual expression values across genes should be centered at zero under above hypothesis. The RLE analysis makes boxplots for all samples based on residual gene expression values, and sorts those boxplots/samples by the interquartile range (IQR) in an increasing order from left to right. Finally, samples at the right most part are likely to be outliers if they have obvious larger IQR than other samples. In practice, we set the right most 5% samples as candidate outliers by default, which could be manually setup in the configuration file.

```{r echo=FALSE, warning=FALSE, results='hide', message=FALSE}
RLEFilterLength <- RLEFilterPercent*ncol(TPM_data)
DSFilter <- DSFilterPercent*ncol(TPM_data)

# Before sample outlier detection and normalization
#RLE
rle=logtpm-apply(logtpm, 1, median) # change "/" to "-" so that we got log(fold-change) which centered on 0 on the RLE plot.
iqr = apply(rle,2,IQR)
rle=melt(cbind(ID=rownames(rle), rle), variable.name = "Sample",value.name ="TPM", id="ID")
names(rle) <- c("feature","Sample","TPM")
rle_IQR <- rle %>% group_by(Sample) %>% summarise(IQR = IQR(TPM))
rle_IQR_range <- rle_IQR$IQR %>% range %>% abs() %>% max()
rle_IQR_range <- 2*rle_IQR_range %>% ceiling()
bymedian <- with(rle, reorder(Sample, TPM, IQR))  # sort by IQR
par(mar=c(3,3,3,3))
boxplot(TPM ~ bymedian, data=rle, outline=F, ylim = c(-rle_IQR_range, rle_IQR_range), las=2, boxwex=1, col='gray', cex.axis=0.3, main="RLE plot before QC", xlab="", ylab="Residual expression levels", frame=F)
ExpPerSample <- nrow(TPM_data)
RLEFilterList <- unique(bymedian[((length(bymedian)-ExpPerSample*RLEFilterLength)+1):length(bymedian)]) #filtered
RLEFilterList <- as.character(RLEFilterList)
print(paste0("The right most ", RLEFilterPercent*100, "% samples (N = ", length(RLEFilterList), ") are marked as candidate outliers in this step:") )
RLEFilterList
```

### Hierarchical clustering plot 

eQTLQC uses (1 - spearman's correlation coefficient) as distance metric to measure sample similarities. Under the assumption that samples should be homogeneous, all samples should have approximate distances between each others, and samples with problematic expression profiles will be far from other samples with normal expression profiles in the clustering dendrogram. To make this process automatic, we use a Mahalanobis distance-based approach to identify outliers. We measure the Mahalanobis distances between each sample and the distribution of all samples. The chi-squared P value was calculated for each sample in each cluster. Clusters with ≤ 60% samples with Bonferroni-corrected P-values ≤ 0.05 will be marked as outlier clusters, and all samples included will be marked as candidate outliers. To efficiently calculate the mahalanobis distance, the top 100 genes sorted by expression variances will be used. All parameters could be manually setup in the configuration file.

Clusters with <= `cluster_percent` (60% in default) samples with Bonferroni-corrected P-values <= `pvalues_cut` (0.05 in default) are marked as outlying clusters, and all samples included will be marked as candidate outliers.

```{r echo=FALSE, warning=FALSE, results='hide', message=FALSE}
#hcluster
sampleDists <- 1 - cor(logtpm, method='spearman')
hc <- hclust(as.dist(sampleDists), method = "complete")
hcphy <- as.phylo(hc)
if(metaInfo && batch_adjust_signal){
 co <- Meta[match(hcphy$tip.label, Meta[[metaInfo_sample_id_column]]), metaInfo_batch_column]
 co <- as.factor(co)
 co_levels <- levels(co)
 co <- brewer.pal(length(co_levels),"Set3")
 par(mar=c(3,3,3,3))
 plot(hcphy, tip.col = co, type = "unrooted", cex=.2, lab4ut='axial',underscore = T, main="Sample clustering before QC (Spearman - Cor.)")
 Xtext =  co_levels #paste("batch",1:co_length)
 legend('bottomleft',pch=21,Xtext, col='white',pt.bg=co, cex=.5)
} else {
  plot(hcphy, type = "unrooted", cex=.2, lab4ut='axial',underscore = T, main="Sample clustering before QC (Spearman - Cor.)")
}

ntop <- topk_genes
Pvars <- apply(logtpm, 1, var)
select <- order(Pvars, decreasing =TRUE)[seq_len(min(ntop, length(Pvars)))]
MD_matrix <- logtpm[select, ]
MahalanobisDistance = mahalanobis(t(MD_matrix), colMeans(t(MD_matrix)), cov(t(MD_matrix))) 
# Note: t(MD_matrix)=sample_row*gene_column, Manhalanobis() returns one vector with length=row number
pvalues = pchisq(MahalanobisDistance, df=nrow(MD_matrix), lower.tail=F)
pvalues.adjust = p.adjust(pvalues, method ="bonferroni") # adjusted pvalues for each sample
pvalues.low <- pvalues.adjust[pvalues.adjust<pvalues.cut]

HCoutliers <- character()
for(x in c(1:treesNum)){
  trees <- cutree(hc,k=x)
  idx <- c(1:x)#which tree is checking
  for(i in idx)
  {
    group <- hc$labels[which(trees == i)]
    if(sum(group %in% names(pvalues.low))/length(group) >= cluster_percent)
    {
      HCoutliers <- union(HCoutliers,group)
    }
  }
}

print(paste(length(HCoutliers), "samples are marked as candidate outlier(s) in this step.", sep = " "))
if(length(HCoutliers)>0){
  print("Sample outliers are marked in red as follows:")
  print(HCoutliers)
  co1 = hc$labels%in%HCoutliers
  co1[which(co1 == "FALSE")]<-"gray0"
  co1[which(co1 == "TRUE")]<-"red"
  par(mar=c(3,3,3,3))
  plot(hcphy, tip.col = co1, type = "unrooted", cex=.2, lab4ut='axial',underscore = T, main="Label Outliers in Red")
  Xcol = c("gray0", "red")
  Xtext = c("Normal Sample", "Outliers")
  legend('bottomleft',pch=21,Xtext, col='white',pt.bg=Xcol, cex=1)
}else{
  print("No outlier detected.")
}
```

### D-statistics distribution
The D-statistic of each sample is defined as the median Spearman’s correlation coefficients with other samples. Samples locating at the left tail of the D-statistic distribution are likely to be outliers. By default, eQTLQC considers the left most 5% samples as candidate outliers. 

```{r,echo=FALSE, warning=FALSE, message=FALSE }
#D-s
D = apply(1-sampleDists, 1, median)
hist(D, breaks=100, ylab="Number of samples", xlab="D-statistic", main="Histogram of Sample D-statistics before data QC")
DSFilter <- sort(D)[DSFilter]
D<-as.data.frame(D)
D<-data.frame(Sample = rownames(D),D = D$D)
D_filterList = D%>%filter(D <= DSFilter)
D_filterList <- D_filterList$Sample
D_filterList<-as.character(D_filterList)
print(paste0("The right most ", DSFilterPercent*100, "% samples (N=", length(D_filterList), ") are marked as candidate outliers in this step:") )
print(D_filterList)
```


### Sample outliers with problematic expression profiles are defined as the intersection of the candidate outliers derived from RLE, H-Clustering and D-statistics distribution analyses.
```{r , echo=FALSE, warning=FALSE, message=FALSE}
outliersList <- c()
outliersList <- intersect(RLEFilterList,intersect(HCoutliers,D_filterList))
print("Outliers:")
outliersList
outliersList<- union(outliersList,clusterFilterList)
```



```{r if no metaInfo, echo=FALSE, warning=FALSE, results='hide', message=FALSE}
if(!metaInfo){
  outliersIndex <- which(colnames(logtpm) %in% outliersList)
  if(!length(outliersIndex) == 0){
    logtpm <- logtpm[,-outliersIndex]
  }

  expr <- logtpm
  # quantile normalization with order preserved. Now TPM is changed to rank normalized gene expression.
  m=apply(expr, 1, mean); 
  sd=apply(expr, 1, sd)
  expr = t(apply(expr, 1, rank, ties.method = "average"));
  #expr = qnorm(expr / (ncol(expr)+1));  # to standard normalization
  expr = qnorm(expr / (ncol(expr)+1), mean=m, sd=sd)  # or, to preserve the mean and sd of each gene
  expr = as.data.frame(expr)
  
  write.table(format(expr, digits=4,nsmall=4), file = "expression.postQC.xls", sep="\t", col.names = NA, quote=F,row.names = TRUE)
}

```


```{r adjust covariates using combat and sva, echo=FALSE, warning=FALSE, results='hide', message=FALSE}
if(metaInfo)
{
  outliersIndex <- which(colnames(logtpm) %in% outliersList)
  if(!length(outliersIndex) == 0){
    logtpm <- logtpm[,-outliersIndex]
  }

  expr <- logtpm
  m=apply(expr, 1, mean); 
  sd=apply(expr, 1, sd)
  expr = t(apply(expr, 1, rank, ties.method = "average"));
  #expr = qnorm(expr / (ncol(expr)+1));  # to standard normalization
  expr = qnorm(expr / (ncol(expr)+1), mean=m, sd=sd)  # or, to preserve the mean and sd of each gene; matrix type
  #expr = as.data.frame(expr)
  covs = Meta
  
  if(gender_adjust_signal)
  {
    	covs[[metaInfo_gender_column]] = as.factor(covs[[metaInfo_gender_column]])
  }
  if(batch_adjust_signal){
      covs[[metaInfo_batch_column]] = as.factor(covs[[metaInfo_batch_column]])
    
      message("# adjusting expression with covariates...")
      ######################
      #Combat remove batch effects
      modcombat = model.matrix(~1, data = covs)
      combat_expr = ComBat(dat=as.matrix(expr), batch=covs[[metaInfo_batch_column]], mod = modcombat)
      # show sample clustering plot after removeing batch effects
      par(mar=c(3,3,3,3))
      sampleDists_combat = 1 - cor(combat_expr, method='spearman')

      hc_combat = hclust(as.dist(sampleDists_combat),method = "complete")
      hcphy_combat = as.phylo(hc_combat)
      co <- Meta[match(hcphy_combat$tip.label, Meta[[metaInfo_sample_id_column]]), metaInfo_batch_column]
      co <- as.factor(co)
      co_levels <- levels(co)
      co <- brewer.pal(length(co_levels),"Set3")

      plot(hcphy_combat, tip.col = co, type = "unrooted", cex=.2, lab4ut='axial',underscore = T, main="Sample clustering after removing batch effects using Combat.")
      Xtext =  co_levels 
      legend('bottomleft',pch=21,Xtext, col='white',pt.bg=co, cex=.5)
    }
    #####################
    #sva adjusted
    covs <- covs %>% select(adjust_covariates) 
    Mod = model.matrix(~. , data=covs) # full model(adjustment variables + variables of interest)
    #Mod = model.matrix(~batch, data=covs)
    Mod0 = model.matrix(~1,data=covs) # Null model (vairables of interest)
  
    svaobj = sva(as.matrix(combat_expr),Mod, Mod0)
    fsvaobj = fsva(dbdat=as.matrix(combat_expr),mod=Mod,sv=svaobj,   newdat=as.matrix(combat_expr))
    residuals = fsvaobj$db
    
  
    message("# run RLE on SVA normalized quantification data ...")
    #####################
    #residuals <- combat_expr
    ## RLE after SVA
    res=data.frame(residuals, check.names = F)
    rle=res-apply(res, 1, median) # after SVA
    rle=melt(cbind(ID=rownames(rle), rle), variable.name = "Sample",value.name ="TPM", id="ID")
    bymedian <- with(rle, reorder(Sample, TPM, IQR))  # sort by IQR
    par(mar=c(3,3,3,3))
    boxplot(TPM ~ bymedian, data=rle, ylim = c(-rle_IQR_range, rle_IQR_range), outline=F, las=2, boxwex=1, col='gray', cex.axis=0.5, main="RLE plot after QC", xlab="", ylab="Residual expression values")
    #abline(h=0, col='red',lwd=1)
    
    ## mean gene expression distribution before and after SVA
    par(mfrow=c(2,1))
    hist(apply(expr,1,mean), breaks=100, xlab="Rank normalized expression log10(TPM)", main="Expression distribution before SVA adjustment")
    hist(apply(residuals,1,mean), breaks=100, xlab="Rank normalized expression log10(TPM)", main="Expression distribution after SVA adjustment")
    ## clustering after sva
    sampleDists_sva = 1 - cor(residuals, method='spearman')
    hc_sva = hclust(as.dist(sampleDists_sva),method = "complete")
    hcphy_sva = as.phylo(hc_sva)
    
    #par(mfrow=c(1,1))
    par(mar=c(3,3,3,3))
    
    co <- Meta[match(hcphy_sva$tip.label, Meta[[metaInfo_sample_id_column]]), metaInfo_batch_column]
    co <- as.factor(co)
    co_levels <- levels(co)
    co <- brewer.pal(length(co_levels),"Set3")
    plot(hcphy_sva, tip.col = co, type = "unrooted", cex=.2, lab4ut='axial',underscore = T, main="Sample clustering after QC.")
    Xtext =  co_levels 
    legend('bottomleft',pch=21,Xtext, col='white',pt.bg=co, cex=.5)
   
  message("# save final quantification data into file")
  ######################
  write.table(format(residuals, digits=4,nsmall=4), file = "expression.postQC.xls", sep="\t", col.names = NA, quote=F,row.names = TRUE)
}

```
